\section{\module{protocols} ---
         Protocol Definition, Declaration, and Adaptation}
\declaremodule{}{protocols}
\moduleauthor{Phillip J. Eby}{pje@telecommunity.com}
\sectionauthor{Phillip J. Eby}{pje@telecommunity.com}
\modulesynopsis{Protocol declaration and adaptation functions as described in
    PEP XXX.}

\begin{quotation}
The typical Python programmer is an integrator, someone who is
connecting components from various vendors.  Often times the
interfaces between these components require an intermediate
adapter.  Usually the burden falls upon the programmer to
study the interface exposed by one component and required by
another, determine if they are directly compatible, or develop
an adapter.  Sometimes a vendor may even include the
appropriate adapter, but then searching for the adapter and
figuring out how to deploy the adapter takes time.

\hfill --- Martelli \& Evans, PEP 246
\end{quotation}

This package builds on the object adaptation protocol presented in \pep{246}
to make it easier for component authors, framework suppliers, and other
developers to:

\begin{itemize}

\item Specify what behavior a component requires or provides

\item Specify how to adapt the interface provided by one component to that
required by another

\item Specify how to adapt objects of a particular type or class (even
built-in types) to a particular required interface

\item Automatically adapt a supplied object to a required interface, and

\item Do all of the above, even when the components or frameworks involved
were not written to take advantage of this package, and even if the frameworks
have different mechanisms for defining interfaces.

\end{itemize}

Assuming that a particular framework either already supports this package, or
has been externally adapted to do so, then framework users will typically
use this package's declaration API to declare what interfaces their classes or
objects provide, and/or to declare adaptations between interfaces or
components.

For framework developers, this package offers an opportunity to replace
tedious and repetitive type-checking code (such as \function{isinstance()},
\function{type()}, \function{hasattr()}, or interface checks) with single
calls to \function{adapt()} instead.  In addition, if the framework has
objects that represent interfaces or protocols, the framework developer can
make them usable with this package's declaration API by adding adapters for
(or direct implementations of) the \class{IOpenProtocol} interface provided
herein.

If the developer of a framework does not do these things, it may still be
possible for a framework user or third-party developer to do them, in order to
be able to use this package's API.  The user of a framework can often call
\function{adapt()} on a component before passing it to a non-adapting
framework.  And, it's possible to externally adapt a
framework's interface objects as well.

For example, the \module{protocols.classic} module includes an adapter that
implements \class{IOpenProtocol} on behalf of Zope \class{Interface}
objects.  This allows them to be used as arguments to this package's protocol
declaration API.  This works even though Zope is completely unaware of the
\module{protocols} package.  (Of course, this does not give Zope
\class{Interface} objects all of the capabilities that \class{Protocol}
objects have, but it does make their existing functionality accessible
through the same API.)

Finally, framework and non-framework developers alike may also wish to use the
\class{Protocol} and \class{Interface} base classes from this package to
define protocols or interfaces of their own, or perhaps use some of the
adaptation mechanisms supplied here to implement ``double dispatching'' or
the ``visitor pattern''.


\begin{seealso}

\seepep{246}{Object Adaptation}{PEP 246 describes an early version of the
adaptation protocol used by this package.}

\end{seealso}



































\subsection{Protocols and Interfaces \label{protocol-concepts}}

Many languages and systems provide ways of defining \strong{interfaces} that
components provide or require.  Some mechanisms are purely for documentation,
others are used at runtime to obtain or verify an implementation.  Typically,
interfaces are formal, intended for compiler-verified static type checking.

As a dynamic language, Python more often uses a looser notion of interface,
known as a \strong{protocol}.  While protocols are often very precisely
specified, their intended audience is a human reader or developer, not a
compiler or automated verification tool.

Automated verification tools, however, usually extract a high overhead cost
from developers.  The Java language, for example, requires that all methods
of an interface be defined by a class that claims to implement the
interface, even if those methods are never used in the program being
compiled!  And yet, the more important \emph{dynamic} behavior of the
interface at runtime is not captured or verifiable by the compiler, so written
documentation for human readers is still required!

In the Python language, the primary uses for objects representing protocols
or interfaces are at runtime, rather than at compile time.  Typically, such
objects are used to ask for an implementation of the interface, or supplied
by an object to claim that it provides an implementation of that interface.

In principle, any Python object may be used as a \strong{protocol object}.
However, for a variety of practical reasons, it is best that protocol objects
be hashable and comparable.  That is, protocol objects should be usable as
dictionary keys.

This still allows for a wide variety of protocol object implementations,
however.  One might assign meaning to the number 42, for example, as
referring to some hypothetical ``hitchhiker'' protocol.  More realistically,
the Microsoft COM framework uses UUIDs (Universally Unique Identifiers) to
identify interfaces.  UUIDs can be represented as Python strings, and thus
are usable as protocol objects.





But a simple string or number is often not very useful as a protocol
object.  Aside from the issue of how to assign strings or numbers to
protocols, these passive protocol objects cannot \emph{do} anything, and by
themselves they document nothing.

There are thus two more common approaches to creating protocol objects in
Python: classes (such as abstract base classes or ``ABCs''), and \strong{
interface objects}.  Interface objects are typically also defined using Python
\code{class} statements, but use a custom metaclass to create an object
that may not be usable in the same ways as a ``real'' Python class.  Many
Python frameworks (such as Twisted, Zope, and this package) provide their own
framework-specific implementations of this ``interface object'' approach.

Since classes and most interface object implementations can be used as
dictionary keys, and because their Python source code can serve as (or
be converted to) useful documentation, both of these approaches are viable
ways to create protocol objects usable with the \module{protocols} package.

In addition, inheriting from a class or interface objects is a simple way to
define implication relationships between protocol objects.  Inheriting from a
protocol to create a new protocol means that the new protocol \strong{implies}
the old protocol.  That is, any implementation or adaptation to the new
protocol, is implied to be usable in a place where the old protocol was
required.  (We will have more to say about direct and adapted implication
relationships later on, in section \ref{proto-implication}.)

At this point, we still haven't described any mechanisms for making adapters
available, or declaring what protocols are supported by a class or object
To do that, we need to define two additional kinds of protocol objects, that
have more specialized abilities.

An \strong{adapting protocol} is a protocol object that is potentially able to
adapt components to support the protocol it represents, or at least to
recognize that a component supports (or claims to support) the protocol.  To
do this, an adapting protocol must have an \method{__adapt__} method, as
will be described in section \ref{adapt-protocol}.  (Often, this method
can be added to an existing class, or patched into an interface object
implementation.)



An \strong{open protocol} is an adapting protocol that is also capable of
accepting adapter declarations, and managing its implication relationships
with other protocols.  Open protocols can be used with this package's
protocol declaration API, as long as they implement (or can be adapted to)
the \class{IOpenProtocol} interface, as will be described in section
\ref{open-protocols}.

Notice that the concepts of protocol objects, adpating protocols, and open
protocols are themselves ``protocols''.  The \module{protocols} package supplies
three interface objects that symbolize these concepts: \class{IProtocol},
\class{IAdaptingProtocol}, and \class{IOpenProtocol}, respectively.  Just as
the English phrases represent the concepts in this text, the interface objects
represent these concepts at runtime.

Whether a protocol object is as simple as a string, or as complex as an
\class{IOpenProtocol}, it can be used to request that a component provide
(or be adaptable to) the protocol that it symbolizes.  In the next section,
we'll look at how to make such a request, and how the different kinds of
protocol objects participate (or not) in fulfilling such requests.






















\subsection{\function{adapt()} and the Adaptation Protocol
\label{adapt-protocol}}

Component adaptation is the central focus of the \module{protocols} package.
All of the package's protocol declaration API depends on component adaptation
in order to function, and the rest of the package is just there to make it
easier for developers to use component adaptation in their frameworks and
programs.

Component adaptation is performed by calling the \function{adapt()} function,
whose design is based largely on the specification presented in \pep{246}:

\begin{funcdesc}{adapt}{component, protocol,
\optional{, default \optional{, factory}}}

Return an implementation of \var{protocol} (a protocol object) for
\var{component} (any object).  The implementation returned may be
\var{component}, or a wrapper that implements the protocol on its
behalf.  If no implementation is available, return \var{default}.  If no
\var{default} is provided, call \code{\var{factory}(\var{component},
\var{protocol})} and return the result.  If no \var{factory} is supplied,
raise \exception{NotImplementedError}.

\var{default} may be supplied as a positional or keyword argument.
\var{factory}, however, must be supplied as a keyword argument in order
to be used.
\end{funcdesc}

The component adaptation process performed by \function{adapt()} proceeds
in four steps:

\begin{enumerate}

\item If the protocol is a class or type, and the component is an instance
of that class or type, the component is returned unchanged.  (This quickly
disposes of the most trivial cases).

\item If the component has a \method{__conform__} method, it is called,
passing in the protocol.  If the method returns a value other than
\constant{None}, it is returned as the result of \function{adapt()}.

\item If the protocol has an \method{__adapt__} method, it is called,
passing in the component.  If the method returns a value other than
\constant{None}, it is returned as the result of \function{adapt()}.

\item Perform default processing as described above, returning \var{default},
invoking \var{factory}, or raising \exception{NotImplementedError} as
appropriate.

\end{enumerate}

This four-step process is called the \strong{adaptation protocol}.  Note
that it can be useful even in the case where neither the component nor the
protocol object are aware that the adaptation protocol exists, and it
gracefully degrades to a kind of \function{isinstance()} check in that
case.  However, if either the component or the protocol object has been
constructed (or altered) so that it has the appropriate \method{__conform__}
or \method{__adapt__} method, then much more meaningful results can be
achieved.

Throughout the rest of this document, we will say that a component
\strong{supports} a protocol, if calling \code{adapt(component,protocol)} does
not raise an error.  That is, a component supports a protocol if its
\method{__conform__} method or the protocol's \method{__adapt__} method
return a non-\constant{None} value.

This is different from saying that an object \strong{provides} a protocol.  An
object provides a protocol if \code{adapt(ob,protocol) is ob}.  Thus,
if an object \emph{provides} a protocol, it \emph{supports} the protocol, but
an object can also support a protocol by having an adapter that provides the
protocol on its behalf.

Now that you know how \function{adapt()} works, you can actually make use of it
without any of the other tools in the \module{protocols} package.  Just define
your own \method{__conform__} and \method{__adapt__} methods, and off
you go!






In practice, however, this is like creating a new kind of Python ``number''
type.  That is, it's certainly possible, but can be rather tedious and is
perhaps best left to a specialist.  For that reason, the \module{protocols}
package supplies some useful basic protocol types, and a ``declaration API''
that lets you declare how protocols, types, and objects should be adapted to
one another.  The rest of this document deals with how to use those types and
APIs.

You don't need to know about those types and APIs to create your own kinds of
protocols or components, just as you don't need to have studied Python's
numeric types or math libraries to create a numeric type of your own.  But,
if you'd like your new types to interoperate well with existing types, and
conform to users' expectations of how such a type behaves, it would be a good
idea to be familiar with existing implementations, such as the ones described
here.


























\subsubsection{Replacing Introspection with Adaptation}

Component adaptation is intended to completely replace all non-cooperative
introspection techniques, such as \function{type()}, \function{isinstance()},
\function{hasattr()}, and even interface checks.  Such introspection
tends to limit framework flexibility by unnecessarily closing policies to
extension by end users.  It often makes code maintenance more difficult as
well, since such checks are often performed in more than one place, and
must be kept in sync whenever a new interface or type must be checked.

The common use cases for such introspection are:

\begin{itemize}

\item To manually adapt a supplied component to a needed interface

\item To select one of several possible behaviors, based on the kind of
component supplied

\item To select another component, or take some action, using information
about the interfaces supported by the supplied component

\end{itemize}

Obviously, the first case is handled quite well by \function{adapt()}, at
least in an environment where it's easy to declare adapters between types and
protocols.  The second and third cases may at first seem to demand an ability
to introspect what interfaces are supported by a component.  But, as we'll
see later, they are almost always better served by defining new protocols
that supply the required behavior or metadata, and then requesting
implementations of those protocols.

In all three use cases, replacing introspection with adaptation opens the
framework to third party extensions, without further modifications being
required -- and without the need to do extensive design or documentation
of a new hook or extension point to be added to the framework.  Indeed,
the availability of a standard mechanism for adaptation means that the
extension mechanism need only be documented once: right here in this
document.


In section \ref{introspect-elim}, we will present examples of how to
refactor all three kinds of introspection code to purely adaptation-driven
code, showing how the flexibility and readability of the code improves in the
process.  But first, we will need to cover how protocols and interfaces can
be defined, declared, and adapted, using the API provided by the
\module{protocols} package.



































\subsubsection{Differences Between \function{protocols.adapt()} and \pep{246}}

If you have read \pep{246} or are looking for an exact implementation of it,
you should know that there are a few differences between the \module{protocols}
implementation of \function{adapt()} and the \pep{246} specification.  If you
don't care about these differences, you can skip this mini-appendix and
proceed directly to section \ref{protocols-defining}, ``Defining and Subclassing
Interfaces''.

The first difference is that \exception{TypeError} is treated differently in
each implementation.  \pep{246} says that if a \method{__conform__} or
\method{__adapt__} method raises a \exception{TypeError}, it should be
treated in the same way as if the method returned \constant{None}.  This was
a workaround for the issue of accidentally calling an unbound class
method, in the case where a component or protocol supplied to
\function{adapt()} was a class.

The \module{protocols} implementation of \function{adapt()} attempts to catch
such errors also, but will reraise any exception that appears to come from
\emph{within} the execution of the \method{__conform__} or
\method{__adapt__} method.  So if these methods raise a \exception{TypeError},
it will be passed through to the caller of \function{adapt}.  Thus, if you
are writing one of these methods, you should not raise a \exception{TypeError}
to signal the lack of an adaptation.  Rather, you should return \constant{None}.

Second, \exception{NotImplementedError} is raised when no adaptation is
found, and no default is supplied, rather than the \exception{TypeError}
specified by \pep{246}.  And third, \function{protocols.adapt()} has an optional
\var{factory} argument that the \pep{246} \function{adapt()} does not.

These differences are the result of experience using the \module{protocols}
package with PEAK, and advances in the Python state-of-the-art since
\pep{246} was written (over two years ago).  We believe that they make the
adaptation protocol more robust, more predictable, and easier to use for
its most common applications.






\subsection{Defining and Subclassing Interfaces \label{protocols-defining}}

The easiest way to define an interface with the \module{protocols} package is
to subclass \class{protocols.Interface}.  \class{Interface} does not supply any
data or methods of its own, so you are free to define whatever you need. There
are two common styles of defining interfaces, illustrated below:

\begin{verbatim%
}from protocols import Interface

# "Pure" interface style

class IReadMapping(Interface):

    """A getitem-only mapping"""

    def __getitem__(key):
        """Return value for key"""


# Abstract Base Class (ABC) style

class AbstractMapping(Interface):

    """A getitem-only mapping"""

    def __getitem__(self,key):
        """Return value for key"""
        raise NotImplementedError
\end{verbatim}

The ``pure'' style emphasizes the interface as seen by the caller, and is not
intended to be subclassed for implementation.  Notice that the \code{self}
parameter is not included in its method definitions, because \code{self} is not
supplied when calling the methods.  The ``ABC'' style, on the other hand,
emphasizes implementation, as it is intended to be subclassed
for that purpose.  Therefore, it includes method bodies, even for abstract
methods.  Each style has different uses: ``ABC'' is a popular rapid development
style, while the ``pure'' approach has some distinct documentation advantages.


\class{protocols.Interface} may be used as a base class for either style.  You
should be aware, however, that Interface objects use an explicit metaclass,
\class{protocols.InterfaceClass}.  This means that if you want to subclass an
interface for implementation using another metaclass, you may need to create a
third metaclass that combines \class{InterfaceClass} with your desired
metaclass.

Subclassing a subclass of \class{Interface} creates a new \class{Interface}
that implies the first interface.  This means that any object that supports
the second interface, is considered to implicitly support the first
interface.  For example:

\begin{verbatim%
}class IReadWriteMapping(IReadMapping):

    """Mapping with getitem and setitem only"""

    def __setitem__(key,value):
        """Store value for key, return None"""

\end{verbatim}

The \code{IReadWriteMapping} interface implies the \code{IReadMapping}
interface.  Therefore, any object that supports \code{IReadWriteMapping} is
understood to also support the \code{IReadMapping} interface.  The reverse,
however, is not true.

Inheritance is only one way to declare that one interface implies another,
however, and its uses are limited.  Let's say for example, that some package
\code{A} supplies objects that support \code{IReadWriteMapping}, while package
\code{B} needs objects that support \code{IReadMapping}.  But each package
declared its own interface, neither inheriting from the other.

As developers reading the documentation of these interfaces, it is obvious to
us that \code{IReadWriteMapping} implies \code{IReadMapping}, because we
understand what they do.  But there is no way for Python to know this, unless
we explicitly state it, like this:




\begin{verbatim%
}import protocols
from A import IReadWriteMapping
from B import IReadMapping

protocols.declareAdapter(
    protocols.NO_ADAPTER_NEEDED,
    provides = [IReadMapping],
    forProtocols = [IReadWriteMapping]
)
\end{verbatim}

In the above example, we use the \module{protocols} declaration API to say that
no adapter is needed to support the \code{B.IReadMapping} interface for
objects that already support the \code{A.IReadWriteMapping} interface.

At this point, if we supply an object that supports \code{IReadWriteMapping},
to a function that expects an \code{IReadMapping}, it should work, as long as
we either call \code{adapt(ob,IReadMapping)} first, or the code we're
calling does so.

There are still other ways to declare that one interface implies another.  For
example, if the author of our example package \code{B} knew about package {A}
and its \code{IReadWriteMapping} interface, he or she might have defined
\code{IReadMapping} this way:
















\begin{verbatim%
}import protocols
from protocols import Interface

from A import IReadWriteMapping

class IReadMapping(Interface):

    """A getitem-only mapping"""

    protocols.advise(
        protocolIsSubsetOf = [IReadWriteMapping]
    )

    def __getitem__(key):
        """Return value for key"""

\end{verbatim}

This is syntax sugar for creating the interface first, and then using
\code{protocols.declareAdapter(NO_ADAPTER_NEEDED)}.  Of course, you can only use
this approach if you are the author of the interface!  Otherwise, you must use
\function{declareAdapter()} after the fact, as in the previous example.

In later sections, we will begin looking at the \module{protocols} declaration
APIs -- like \function{declareAdapter()} and \function{advise()} -- in more
detail.  But first, we must look briefly at the interfaces that the
\module{protocols} package expects from the protocols, adapters, and other
objects supplied as parameters to the declaration API.












\subsection{Interfaces Used by the Declaration API}

Like any other API, the \module{protocols} declaration API has certain
expectations regarding its parameters.  These expectations are documented and
referenced in code using interfaces defined in the \module{protocols.interfaces}
module.  (The interfaces are also exported directly from the top level of the
\module{protocols} package.)

You will rarely use or subclass any of these interface objects, unless you are
customizing or extending the system.  Four of the interfaces exist exclusively
for documentation purposes, while the rest are used in \function{adapt()} calls
made by the API.

First, let's look at the documentation-only interfaces.  It is not necessary
for you to declare that an object supports these interfaces, and the
\module{protocols} package never tries to \function{adapt()} objects to them.

\begin{description}

\item[IAdapterFactory] \hfill \\
Up until this point, we've been talking about ``adapters'' rather loosely.  The
\class{IAdapterFactory} interface formalizes the concept.  An \strong{adapter
factory} is a callable object that takes two arguments (an object and a
protocol, in that order) and returns an object that provides the protocol on
behalf of the passed-in object.

You'll notice that this is the same signature required of the optional
\var{factory} argument to \function{adapt()}.  It is also the signature required
of objects supplied as ``adapter'' or ``factory'' arguments to the declaration
API functions.

The \module{protocols} package supplies two functions that provide
this interface: \function{NO_ADAPTER_NEEDED} and \function{DOES_NOT_SUPPORT}.
\function{NO_ADAPTER_NEEDED} is used to declare that an object provides a
protocol directly, and thus it returns the object passed into it, rather than
some kind of adapter.  \function{DOES_NOT_SUPPORT} is used to declare that an
object does not support a protocol, even with an adapter.  (Since this is the
default case, \function{DOES_NOT_SUPPORT} is rarely used, except to indicate
that a subclass does not support an interface that one of its superclasses
does.)

\item[IProtocol] \hfill \\
This interface formalizes the idea of a ``protocol object''.  As you will
recall from section \ref{protocol-concepts}, a protocol object is any object
that can be used as a Python dictionary key.  The second argument to
\function{adapt()} must be a protocol object according to this definition.


\item[IAdaptingProtocol] \hfill \\
This interface formalizes the idea of an ``adapting protocol'', specifically
that it is a protocol object (i.e., it provides \class{IProtocol}) that also
has an \method{__adapt__} method as described in section \ref{adapt-protocol}.
\class{IAdaptingProtocol} is a subclass of \class{IProtocol}, so of course
\function{adapt()} accepts such objects as protocols.

\item[IImplicationListener] \hfill \\
This interface is for objects that want to receive notification when new
implication relationships (i.e. adapters) are registered between two protocols.
If you have objects that want to keep track of what interfaces they support,
you may want those object to implement this interface so they can be kept
informed of new protocol-to-protocol adapters.

\end{description}

The other three interfaces are critical to the operation of the declaration API,
and thus must be supported by objects supplied to it.  The \module{protocols}
package supplies and registers various adapter classes that provide these
interfaces on behalf of many commonly used Python object types.  So, for each
interface, we will list ``known supporters'' of that interface, whether they
are classes supplied by \module{protocols}, or built-in types that are
automatically adapted to the interface.

We will not, however, go into details here about the methods and behavior
required by each interface.  We will leave such detailed discussions to
section \ref{customizing-adaptation}, ``Customizing and Extending the
Declaration API'', since customization and extension are all that the
details are useful for.

\begin{description}



\item[IOpenProtocol] \hfill \\
This interface formalizes the ``open protocol'' concept that was introduced
in section \ref{protocol-concepts}.  An \class{IOpenProtocol} is an
\class{IAdaptingProtocol} that can also accept declarations made by the
\module{protocols} declaration API.

The \module{protocols} package supplies two implementations of this interface:
\class{Protocol} and \class{InterfaceClass}.  Thus, any \class{Interface}
subclass or \class{Protocol} instance is automatically considered to provide
\class{IOpenProtocol}.  \note{\class{Interface} is an instance of
\class{InterfaceClass}, and thus provides \class{IOpenProtocol}.  But if you
create an instance of an \class{Interface}, that object does not provide
\class{IOpenProtocol}, because the interfaces provided by an object and its
class (or its instances) can be different.}

In addition to its built-in implementations, the \module{protocols} package
also supplies and declares an adapter factory that adapts Zope X3's
interface objects to the \class{IOpenProtocol} interface, thus allowing
you to use Zope interfaces in calls to the declaration API.  Similar adapters
for other frameworks' interfaces may be added, if there is sufficient demand
and/or contributed code, and the frameworks' authors do not add the adapters
to their frameworks.


\item[IOpenImplementor] \hfill \\
An \class{IOpenImplementor} is a class or type that can be told (via the
declaration API) what protocols its instances provide (or support via an
\class{IAdapterFactory}).  Note that this implies that the instances have
a \method{__conform__} method, or else they would not be able to tell
\function{adapt()} about the declared support!

Support for this interface is optional, since types that don't support it
can still have their instances be adapted by \class{IOpenProtocol} objects.
The \module{protocols} package does not supply any implementations or adapters
for this interface, either.  It is intended primarily as a hook for classes
to be able to receive notification about protocol declarations for their
instances.




\item[IOpenProvider] \hfill \\
Because objects' behavior usually comes from a class definition, it's not that
often that you will declare that a specific object provides or supports an
interface.  But objects like functions and modules do not have a class
definition, and classes themselves sometimes provide an interface.  (For
example, one could say that class objects provide an \class{IClass} interface.)
So, the declaration API needs to also be able to declare what protocols an
individual object (such as a function, module, or class) supports or provides.

That's what the \class{IOpenProvider} interface is for.  An
\class{IOpenProvider} is an object with a \class{__conform__} method, that can
be told (via the declaration API) what protocols it provides (or supports via
an \class{IAdapterFactory}).

Notice that this is different from \class{IOpenImplementor}, which deals with
an class or type's instances.  \class{IOpenProvider} deals with the object
itself.  A single object can potentially be both an \class{IOpenProvider} and an
\class{IOpenImplementor}, if it is a class or type.

The \module{protocols} package supplies and declares an adapter factory that
can adapt most Python objects to support this interface, assuming that they
have a \code{__dict__} attribute.  Thus, it is acceptable to pass a Python
function, module, or instance of a  ``classic'' class to any declaration API
that expects an \class{IOpenProvider} argument.

We'll talk more about making protocol declarations for individual objects
(as opposed to types) in section \ref{protocols-instances}, ``Protocol
Declarations for Individual Objects''.

\end{description}











\subsection{Declarating Implementations and Adapters}

There are three types of relationships that a protocol can participate in:

\begin{itemize}
\item A relationship from a class or type, to a protocol its instances provide
or adapt to,

\item A relationship from an instance, to a protocol it provides or adapts to,
and

\item A relationship from a protocol, to a protocol that it implies or is
adaptable to.
\end{itemize}

All of these relationships may be declared by reference to the source type,
instance or protocol, the desired destination protocol, and an adapter factory
to be used. (In the event that no adapter is needed, the special adapter
factory \function{NO_ADAPTER_NEEDED} is specified.)

To declare relationships like these, the \module{protocols} declaration API
provides three ``primitive'' declaration functions.  Each takes a \var{protocol}
argument that must support the \class{IOpenProtocol} interface, an
adapter function (or \function{NO_ADAPTER_NEEDED}), and the source
type, instance, or protocol.  The functions are
\function{declareAdapterForType()}, \function{declareAdapterForObject()}, and
\function{declareAdapterForProtocol()}, respectively.

You will not ordinarily use these primitives, however, unless you are
customizing or extending the framework.  It is generally easier to call one
of the higher level functions in the declaration API.  These higher-level
functions may make several calls to the primitive functions on your behalf, or
supply useful defaults for certain parameters.  They are, however, based
entirely on the primitive functions, which is important for customizations and
extensions.  This will be covered more in section \ref{customizing-adaptation}.

The next higher layer of declaration APIs are the explicit declaration
functions, \function{declareImplementation}, \function{declareAdapter}, and
\function{adviseObject}.  These functions are structured to support the most
common declaration use cases.

For declaring protocols related to a type or class:

\begin{funcdesc}{declareImplementation}{typ
\optional{, instancesProvide=[ ]} \optional{, instancesDoNotProvide=[ ]}}

Declare that instances of class or type \var{typ} do or do not provide
implementations of the specified protocols.  \var{instancesProvide} and
\var{instancesDoNotProvide} must be sequences of protocol objects that
provide (or are adaptable to) the \class{IOpenProtocol} interface,
such as \class{protocols.Interface} subclasses, or Zope \class{Interface}
objects.

This function is shorthand for calling \function{declareAdapterForType()}
with \function{NO_ADAPTER_NEEDED} and \function{DOES_NOT_SUPPORT} as adapters
from the type to each of the specified protocols.  Note, therefore, that the
listed protocols must be adaptable to \class{IOpenProtocol}.  See
\function{declareAdapterForType()} in section \ref{protocols-contents} for
details.
\end{funcdesc}


For declaring protocols related to a specific, individual instance:

\begin{funcdesc}{adviseObject}{ob
\optional{, provides=[ ]} \optional{, doesNotProvide=[ ]}}
Declare that \var{ob} provides (or does not provide) the specified protocols.
This is shorthand for calling \function{declareAdapterForObject()}
with \function{NO_ADAPTER_NEEDED} and \function{DOES_NOT_SUPPORT} as adapters
from the object to each of the specified protocols.  Note, therefore, that
\var{ob} may need to support \class{IOpenProvider}, and the listed protocols
must be adaptable to \class{IOpenProtocol}.  See
\function{declareAdapterForObject()} in section \ref{protocols-contents} for
details.  Also, see section \ref{protocols-instances}, ``Protocol Declarations
for Individual Objects'', for more information on using
\function{adviseObject}.
\end{funcdesc}





And for declaring all other kinds of protocol relationships:

\begin{funcdesc}{declareAdapter}{factory, provides,
\optional{, forTypes=[ ]} \optional{, forProtocols=[ ]}
\optional{, forObjects=[ ]}}

Declare that \var{factory} is an \class{IAdapterFactory} whose return value
provides the protocols listed in \var{provides} as an adapter for the
classes/types listed in \var{forTypes}, for objects providing the protocols
listed in \var{forProtocols}, and for the specific objects listed in
\var{forObjects}.

This function is shorthand for calling the primitive declaration
functions for each of the protocols listed in \var{provides} and each of the
sources listed in the respective keyword arguments.
\end{funcdesc}

Although these forms are easier to use than raw \code{declareAdapterForX}
calls, they still require explicit reference to the types or objects involved.
For the most common use cases, such as declaring protocol relationships to a
class, or declaring an adapter class, it is usually easier to use the ``magic''
\function{protocols.advise()} function, which we will discuss next.



















\subsubsection{Convenience Declarations in Class, Interface and Module Bodies \label{protcols-advise}}

Adapters, interfaces, and protocol implementations are usually defined in
Python \code{class} statements.  To make it more convenient to make protocol
declarations for these classes, the \module{protocols} package supplies the
\function{advise()} function.  This function  can make declarations about a
class, simply by being called from the body of that class.  It can also be
called from the body of a module, to make a declaration about the module.

\begin{funcdesc}{advise}{**kw}
Declare protocol relationships for the containing class or module.  All
parameters must be supplied as keyword arguments.  This function must be
called directly from a class or module body, or a \exception{SyntaxError}
results at runtime.  Different arguments are accepted, according to whether
the function is called within a class or module.

When invoked in the top-level code of a module, this function only accepts
the \code{moduleProvides} keyword argument.  When invoked in the body of a
class definition, this function accepts any keyword arguments \emph{except}
\code{moduleProvides}.  The complete list of keyword arguments follows.
Unless otherwise specified, protocols must support the \class{IOpenProtocol}
interface.

\note{When used in a class body, this function works by temporarily replacing
the \code{__metaclass__} of the class.  If your class sets an explicit
\code{__metaclass__}, it must do so \emph{before} \function{advise()} is
called, or the protocol declarations will not occur!}
\end{funcdesc}













Keyword arguments accepted by \function{advise()}:

\begin{description}

\item[instancesProvide = \var{protocols}] \hfill \\
A sequence of protocols that instances of the containing class provide, without
needing an adapter.  Supplying this argument is equivalent to calling
\code{declareImplementation(\var{containing class},\var{protocols})}.

\item[instancesDoNotProvide = \var{protocols}] \hfill \\
A sequence of protocols that instances of the containing class do not provide.
This is primarily intended for "rejecting" protocols provided or supported
by base classes of the containing class.  Supplying this argument is equivalent
to calling \code{declareImplementation(\var{containing
class},instancesDoNotProvide=\var{protocols})}.

\item[asAdapterForTypes = \var{types}] \hfill \\
Declare the containing class as an adapter for \var{types}, to the protocols
listed by the \code{instancesProvide} argument (which must also be supplied).
Supplying this argument is equivalent to calling
\code{declareAdapter(\var{containing class}, \var{instancesProvide},
forTypes=\var{types})}.  (Note that this means the containing class must be an
object that provides \class{IAdapterFactory}; i.e., its constructor should
accept being called with two arguments: the object to be adapted, and the
target protocol.)

\item[asAdapterForProtocols = \var{protocols}] \hfill \\
Declare the containing class as an adapter for \var{protocols}, to the
protocols listed by the \code{instancesProvide} argument (which must also be
supplied).  Supplying this argument is equivalent to calling
\code{declareAdapter(\var{containing class}, \var{instancesProvide},
forProtocols=\var{types})}.  (Note that this means the containing class must be
an object that provides \class{IAdapterFactory}; i.e., its constructor should
accept being called with two arguments: the object to be adapted, and the
target protocol.)






\item[protocolExtends = \var{protocols}] \hfill \\
Declare that the containing class is a protocol that extends (i.e., implies)
the listed protocols.  This keyword argument is intended for use inside class
statements that themselves define protocols, such as \class{Interface}
subclasses, and that need to ``inherit'' from incompatible protocols.  For
example, an \class{Interface} cannot directly subclass a Zope interface,
because their metaclasses are incompatible.  But using \code{protocolExtends}
works around this:

\begin{verbatim%
}import protocols
from mypackage import ISomeInterface
from zope.something.interfaces import ISomeZopeInterface

class IAnotherInterface(ISomeInterface):
    protocols.advise(
        protocolExtends = [ISomeZopeInterface]
    )
    #... etc.
\end{verbatim}

In the above example, \code{IAnotherInterface} wants to extend both
\code{ISomeInterface} and \code{ISomeZopeInterface}, but cannot do so directly
because the interfaces are of incompatible types.  \code{protocolExtends}
informs the newly created interface that it implies \code{ISomeZopeInterface},
even though it isn't derived from it.

Using this keyword argument is equivalent to calling
\code{declareAdapter(NO_ADAPTER_NEEDED, \var{protocols},
forProtocols=[\var{containing class}])}.  Note that this means that the
containing class must be an object that supports \class{IOpenProtocol}, such
as an \class{Interface} subclass.









\item[protocolIsSubsetOf = \var{protocols}] \hfill \\
Declare that the containing class is a protocol that is implied (extended) by
the listed protocols.  This is just like \code{protocolExtends}, but in the
``opposite direction''.  It allows you to declare (in effect) that some other
interface is actually a subclass of (extends, implies) this one.  See the
examples in section \ref{protocols-defining} for illustration.

Using this keyword argument is equivalent to calling
\code{declareAdapter(NO_ADAPTER_NEEDED, [\var{containing class}],
forProtocols=\var{protocols})}.

\item[classProvides = \var{protocols}] \hfill \\
Declare that the containing class \emph{itself} provides the specified
protocols.  Supplying this argument is equivalent to calling
\code{adviseObject(\var{containing class}, \var{protocols})}.  Note that this
means that the containing class must support the \class{IOpenProvider}
interface.  The \module{protocols} package supplies default adapters to
support \class{IOpenProvider} for both classic and new-style classes, as long
as they do not have custom \method{__conform__} methods.  See section
\ref{protocols-instances}, ``Protocol Declarations for Individual Objects''
for more details.

\item[classDoesNotProvide = \var{protocols}] \hfill \\
Declare that the containing class \emph{itself} does not provide the specified
protocols.  This is for classes that need to reject inherited class-level
\code{classProvides} declarations.  Supplying this argument is equivalent to
calling \code{adviseObject(\var{containing class},
doesNotProvide=\var{protocols})}, and the \class{IOpenProvider} requirements
mentioned above for \code{classProvides} apply here as well.

\item[moduleProvides = \var{protocols}] (module context only) \hfill \\
A sequence of protocols that the enclosing module provides.  Equivalent to
\code{adviseObject(\var{containing module}, \var{protocols})}.

\end{description}






\subsection{Creating and Using Adapters, Components, and Protocols}

IAdapterFactory

Defining __conform__ - interactions?

Rejecting inherited contracts

object identity; adaptation reflexivity or lack thereof
































\subsection{Protocol Declarations for Individual Objects \label{protocols-instances}}

Because objects' behavior usually comes from a class definition, it's not too
often that you will declare that an individual object provides or supports an
interface, as opposed to making a blanket declaration about an entire class or
type of object.  But objects like functions and modules do not \emph{have} a
class definition that encompasses their behavior, and classes themselves
sometimes provide an interface (e.g. via \function{classmethod} objects).

So, the declaration API needs to also be able to declare what protocols an
individual object (such as a function, module, or class) supports or provides.
This is what \function{adviseObject()} and the
\code{classProvides}/\code{classDoesNotProvide} keywords of
\function{advise()} do.

In most cases, for an object to be usable with \function{adviseObject()}, it
must support the \class{IOpenProvider} interface.  Since many of the objects
one might wish to use with \function{adviseObject()} (such as modules,
functions, and classes) do not directly provide this interface, the
\module{protocols.classic} module supplies and declares an adapter factory tha
can adapt most Python objects to support this interface, assuming that they
have a \code{__dict__} attribute.

This default adapter works well for many situations, but it has some
limitations you may need to be aware of.  First, it works by ``poking'' a new
\method{__conform__} method into the adapted object.  If the object already
has a \method{__conform__} method, a \exception{TypeError} will be raised.  So,
if you need an object to be an \class{IOpenProvider}, but it has a
\method{__conform__} method, you may want to have its class include
\class{ProviderMixin} among its base classes, so that your objects won't rely
on the default adapter for \class{IOpenProvider}.  (See \class{ProviderMixin}
in section \ref{protocols-contents} for more on this.)

Both the default adapter and \class{ProviderMixin} support inheritance of
protocol declarations, when the object being adapted is a class or type.  In
this way, \code{advise(classProvides=\var{protocols})} declarations (or
\code{adviseObject(someClass,\var{protocols})} calls) are inherited by
subclasses.  Of course, you can always reject inherited protocol information
using \code{advise(classDoesNotProvide=\var{protocols})} or
\function{adviseObject(newClass,doesNotProvide=\var{protocols})}.

Both the default adapter and \class{ProviderMixin} work by keeping a mapping of
protocols to adapter factories.  Keep in mind that this means the protocols and
adapter factories will continue to live until your object is garbage collected.
Also, that means for your object to be pickleable, all of the protocols and
adapter factories used must be pickleable.  (This latter requirement can be
quite difficult to meet, since adapter factories are often functions.)

XXX should we add __reduce__ to NO_ADAPTER_NEEDED/DOES_NOT_SUPPORT?  NB
composed adapters aren't currently pickleable either.

Note that none of these restrictions apply if you are only using declarations
about types and protocols, as opposed to individual objects.  (Or if you only
make individual-object declarations for functions, modules, and classes.)
Also note that if you have some objects that need to dynamically support or
not support a protocol on a per-instance basis, then \function{adviseObject()}
is probably not what you want anyway!  Instead, give your objects' class a
\method{__conform__()} method that does the right thing when the object is
asked to conform to a protocol.  \function{adviseObject()} is really intended
for adding metadata to objects that ``don't know any better''.

In general, protocol declarations are a \emph{static} mechanism: they cannot be
changed or removed at will, only successively refined.  All protocol
declarations made must be consistent with the declarations that have already
been made.  This makes them unsuitable as a mechanism for dynamic behavior such
as supporting a protocol based on an object's current state.

In the next section, we'll look more at the static nature of declarations,
and explore what it means to make conflicting (or refining) protocol
declarations.

XXX This section seems awfully short; wasn't there more to say about this?
Maybe it should be a subsubsection.









\subsection{Protocol Implication and Adapter Precedence \label{proto-implication}}

transitivity

adapters can be composed

``adapter path''

``shortest path''

overriding existing declarations

ambiguous override

how circularity is avoided

Use of subsetting to create ``context'' of sorts

GC consequences of implication: protocols must hold references to implied
protocols, so subsetting means subset protocol will exist as long as the
superset protocol does.




















\subsection{Package Contents and Contained Modules\label{protocols-contents}}

The following functions, classes, and interfaces are available from the
top-level \module{protocols} package.  (The list is arranged in order of
their expected frequency of use, with most frequently used items listed first.)

\begin{funcdesc}{adapt}{component, protocol,
\optional{, default \optional{, factory}}}

Return an implementation of \var{protocol} (a protocol object) for
\var{component} (any object).  The implementation returned may be
\var{component}, or an adapter that implements the protocol on its
behalf.  If no implementation is available, return \var{default}.  If no
\var{default} is provided, call \code{\var{factory}(\var{component},
\var{protocol})} and return the result.  If no \var{factory} is supplied,
raise \exception{NotImplementedError}.

\var{default} may be supplied as a positional or keyword argument.
\var{factory}, however, must be supplied as a keyword argument in order
to be used.

A detailed description of this function's operations and purpose may be found in
section \ref{adapt-protocol}.
\end{funcdesc}

\begin{classdesc*}{Interface}
Subclass this to create an interface.  See section \ref{protocols-defining}
for more details.
\end{classdesc*}












\begin{funcdesc}{advise}{**kw}
Declare protocol relationships for the containing class or module.  All
parameters must be supplied as keyword arguments.  This function must be
called directly from a class or module body, or a \exception{SyntaxError}
results at runtime.  Different arguments are accepted, according to whether
the function is called within a class or module.

When invoked in the top-level code of a module, this function only accepts
the \code{moduleProvides} keyword argument.  When invoked in the body of a
class definition, this function accepts any keyword arguments \emph{except}
\code{moduleProvides}.  The complete list of keyword arguments can be found
in section \ref{protcols-advise}.

\note{When used in a class body, this function works by temporarily replacing
the \code{__metaclass__} of the class.  If your class sets an explicit
\code{__metaclass__}, it must do so \emph{before} \function{advise()} is
called, or the protocol declarations will not occur!}
\end{funcdesc}


\begin{funcdesc}{NO_ADAPTER_NEEDED}{component, protocol}
This function simply returns \var{component}.  It is a placeholder used whenever
an object, type, or protocol directly implements or implies another protocol.
Whenever an adapter is not required, but the \module{protocols} API function
you are calling requires an adapter, you should supply this function as the
adapter.  Some protocol implementations, such as the one for Zope interfaces,
are unable to handle adapters other than \function{NO_ADAPTER_NEEDED} and
\function{DOES_NOT_SUPPORT}.
\end{funcdesc}


\begin{funcdesc}{DOES_NOT_SUPPORT}{component, protocol}
This function simply returns \constant{None}.  It is a placeholder used whenever
an object, type, or protocol does not implement or imply another protocol.
Whenever adaptation is not possible, but the \module{protocols} API function
you are calling requires an adapter, you should supply this function as the
adapter.  Some protocol implementations, such as the one for Zope interfaces,
are unable to handle adapters other than \function{NO_ADAPTER_NEEDED} and
\function{DOES_NOT_SUPPORT}.
\end{funcdesc}

\begin{funcdesc}{declareImplementation}{typ
\optional{, instancesProvide=[ ]} \optional{, instancesDoNotProvide=[ ]}}

Declare that instances of class or type \var{typ} do or do not provide
implementations of the specified protocols.  \var{instancesProvide} and
\var{instancesDoNotProvide} must be sequences of protocol objects that
provide (or are adaptable to) the \class{IOpenProtocol} interface,
such as \class{protocols.Interface} subclasses, or Zope \class{Interface}
objects.

This function is shorthand for calling \function{declareAdapterForType()}
with \function{NO_ADAPTER_NEEDED} and \function{DOES_NOT_SUPPORT} as adapters
from the type to each of the specified protocols.  Note, therefore, that the
listed protocols must be adaptable to \class{IOpenProtocol}.
\end{funcdesc}


\begin{funcdesc}{adviseObject}{ob
\optional{, provides=[ ]} \optional{, doesNotProvide=[ ]}}
Declare that \var{ob} provides (or does not provide) the specified protocols.
This is shorthand for calling \function{declareAdapterForObject()}
with \function{NO_ADAPTER_NEEDED} and \function{DOES_NOT_SUPPORT} as adapters
from the object to each of the specified protocols.  Note, therefore, that
\var{ob} may need to support \class{IOpenProvider}, and the listed protocols
must be adaptable to \class{IOpenProtocol}.  See section
\ref{protocols-instances}, ``Protocol Declarations for Individual Objects'',
for more information on using \function{adviseObject}.
\end{funcdesc}













\begin{funcdesc}{declareAdapter}{factory, provides
\optional{, forTypes=[ ]} \optional{, forProtocols=[ ]}
\optional{, forObjects=[ ]}}

Declare that \var{factory} is an \class{IAdapterFactory} whose return value
provides the protocols listed in \var{provides} as an adapter for the
classes/types listed in \var{forTypes}, for objects providing the protocols
listed in \var{forProtocols}, and for the specific objects listed in
\var{forObjects}.

This function is shorthand for calling the primitive declaration
functions (\function{declareAdapterForType},
\function{declareAdapterForProtocol}, and \function{declareAdapterForObject})
for each of the protocols listed in \var{provides} and each of the
items listed in the respective keyword arguments.
\end{funcdesc}

























\begin{classdesc*}{Protocol}
\class{Protocol} is a base class that implements the \class{IOpenProtocol}
interface, supplying internal adapter registries for adapting from other
protocols or types/classes.  Note that you do not necessarily need to use this
class (or any other \class{IOpenProtocol} implementation) in
your programs.  Any object that implements the simpler \class{IProtocol} or
\class{IAdaptingProtocol} interfaces may be used as protocols for the
\function{adapt()} function.  Compliance with the \class{IOpenProtocol}
interface is only required to use the \module{protocols} declaration API.
(That is, functions whose names begin with \code{declare} or \code{advise}.)

To create protocols dynamically, you can create individual \class{Protocol}
instances, and then use them with the declaration API.  You can also subclass
\class{Protocol} to create your own protocol types.  If you override
\function{__init__}, however, be sure to call \function{Protocol.__init__()}
in your subclass' \function{__init__} method.
\end{classdesc*}


\begin{classdesc}{InterfaceClass}{name, bases, dictionary}
\class{InterfaceClass}, a subclass of \class{Protocol} and \class{type}, is a
metaclass used to create new interfaces (i.e., protocol objects) using class
statements.  You may set a class' \code{__metaclass__} to
\class{InterfaceClass}, or simply subclass \class{Interface}.  Normally, you
will only use \class{InterfaceClass} if you need to combine it with another
metaclass, in a subclass of \class{Interface}.
\end{classdesc}














\begin{classdesc*}{ProviderMixin}
If you have a class with a \method{__conform__} method for its instances, but
you also want the instances to support \class{IOpenprovider} (so that
\function{adviseObject} can be used on them), you may want to include this
class as one of your class' bases.  The default adapters for
\class{IOpenprovider} can only adapt objects that do not already have a
\method{__conform__} method of their own.

So, to support \class{IOpenprovider} with a custom \method{__conform__}
method, subclass \class{ProviderMixin}, and have your \method{__conform__}
method invoke the base \method{__conform__} method as a default, using
\function{supermeta()}.  (E.g. \code{return
supermeta(MyClass,self).__conform__(protocol)}.)  See below for more on
the \function{supermeta()} function.
\end{classdesc*}


\begin{funcdesc}{metamethod}{func}
Wrap \var{func} in a manner analagous to \function{classmethod} or
\function{staticmethod}, but as a metaclass-level method that may be redefined
by metaclass instances for their instances.  For example, if a metaclass wants
to define a \method{__conform__} method for its instances (i.e. classes), and
those instances (classes) want to define a \method{__conform__} method for
\emph{their} instances, the metaclass should wrap its \method{__conform__}
method with \function{metamethod}.  Otherwise, the metaclass'
\method{__conform__} method will be hidden by the class-level
\method{__conform__} defined for the class' instances.
\end{funcdesc}


\begin{funcdesc}{supermeta}{typ, ob}
Emulates the Python built-in \function{super()} function, but with support for
metamethods.  If you ordinarily would use \function{super()}, but are calling a
\function{metamethod}, you should use \function{supermeta()} instead.  This is
because Python 2.2 does not support using super with properties (which is
effectively what metamethods are).

Note that if you are subclassing \class{ProviderMixin}
\end{funcdesc}


\begin{funcdesc}{declareAdapterForType}{protocol, adapter, typ \optional{,
depth=1}}
Declare that \var{adapter} adapts instances of class or type \var{typ}
to \var{protocol}, by adapting \var{protocol} to \class{IOpenProtocol} and
calling its \function{registerImplementation} method.  If \var{typ} is adaptable
to \class{IOpenImplementor}, its \function{declareClassImplements} method is
called as well.
\end{funcdesc}


\begin{funcdesc}{declareAdapterForObject}{protocol, adapter, ob \optional{,
depth=1}}
Declare that \var{adapter} adapts the object \var{ob} to \var{protocol}, by
adapting \var{protocol} to \class{IOpenProtocol} and calling its
\function{registerObject} method.  Typically, \var{ob} must support
\class{IOpenProvider}.  See section \ref{protocols-instances} for details.
\end{funcdesc}


\begin{funcdesc}{declareAdapterForProtocol}{protocol, adapter, proto \optional{,
depth=1}}
Declare that \var{adapter} adapts objects that provide protocol \var{proto}
to \var{protocol}, by adapting \var{protocol} to \class{IOpenProtocol} and
calling its \function{addImpliedProtocol} method.
\end{funcdesc}
















\begin{classdesc*}{IOpenProtocol}
This interface documents the behavior required of protocol objects in order to
be used with the \module{protocols} declaration API (the functions whose names
begin with \code{declare} or \code{advise}.)  The declaration API functions
will attempt to \function{adapt()} supplied protocols to this interface.

The methods an \class{IOpenProtocol} implementation must supply are:


\begin{methoddesc}{addImpliedProtocol}{proto, adapter, depth}
Declare that this protocol can be adapted to protocol \var{proto} via the
\class{IAdapterFactory} supplied in \var{adapter}, at the specified implication
level \var{depth}.  The protocol object should ensure that the implied protocol
is able to adapt objects implementing its protocol (typically by recursively
invoking \function{declareAdapterForType()} with increased depth and
appropriately composed adapters), and notify any registered implication
listeners via their \method{newProtocolImplied()} methods.  If the protocol
already implied \var{proto}, this method should have no effect and send no
notifications unless the new \var{adapter} and \var{depth} represent a
``shorter path'' as described in section \ref{proto-implication}.
\end{methoddesc}


\begin{methoddesc}{registerImplementation}{klass, adapter, depth}
Declare that instances of type or class \var{klass} can be adapted to this
protocol via the \class{IAdapterFactory} supplied in \var{adapter}, at the
specified implication level \var{depth}.  Unless \var{adapter} is
\function{DOES_NOT_SUPPORT}, the protocol object must ensure that
any protocols it implies are also able to perform the adaptation (typically
by recursively invoking \function{declareAdapterForType()} with increased depth
and appropriately composed adapters for its implied protocols).  If the
protocol already knew a way to adapt instances of \var{klass}, this method
should be a no-op unless the new \var{adapter} and \var{depth} represent a
``shorter path'' as described in section \ref{proto-implication}.
\end{methoddesc}






\begin{methoddesc}{registerObject}{ob, adapter, depth}
Ensure that the specific object \var{ob} will be adapted to this protocol via
the \class{IAdapterFactory} supplied in \var{adapter}, at the specified
implication level \var{depth}.  The protocol object must also ensure that the
object can be adapted to any protocols it implies.  This method may be
implemented by adapting \var{ob} to \class{IOpenProvider}, calling the
\method{declareProvides()} method, and then recursively invoking
\function{declareAdapterForObject} with increased depth and appropriately
composed adapters for the protocols' implied protocols.
\end{methoddesc}

\begin{methoddesc}{addImplicationListener}{listener}
Ensure that \var{listener} (an \class{IImplicationListener}) will be notified
whenever an implied protocol is added to this protocol, or an implication path
from this protocol is shortened.  The protocol should at most retain a weak
reference to \var{listener}.  Note that if a protocol can guarantee that no
notices will ever need to be sent, it is free to implement this method as a
no-op.  For example, Zope interfaces cannot imply any protocols besides their
base interfaces, which are not allowed to change.  Therefore, no change
notifications would ever need to be sent, so the \class{IOpenProtocol} adapter
for Zope interfaces implements this method as a no-op.
\end{methoddesc}


\note{\class{IOpenProtocol} is a subclass of \class{IAdaptingProtocol}, which
means that implementations must therefore meet its requirements as well, such
as having an \method{__adapt__()} method.}

\end{classdesc*}












\begin{classdesc*}{IOpenProvider}
This interface documents the behavior required of an object to be usable with
\function{adviseObject()}.  Note that some protocol objects, such as the
\class{IOpenProtocol} adapter for Zope interfaces, can handle
\function{adviseObject()} operations without adpating the target object to
\class{IOpenProvider}.  This should be considered an exception, rather than the
rule.  However, the \module{protocols} package declares default adapters so
that virtually any Python object that doesn't already have a
\method{__conform__()} method can be adapted to \class{IOpenProvider}
automatically.

\begin{methoddesc}{declareProvides}{protocol, adapter, depth}
Declare that this object can provide \var{protocol} if adapted by the
\class{IAdapterFactory} supplied in \var{adapter}, at implication level
\var{depth}.  If the object already knew a way to adapt to \var{protocol}, this
method should be a no-op unless the new \var{adapter} and \var{depth} represent
a ``shorter path'' as described in section \ref{proto-implication}.  Typically,
an implementation of this method will also adapt \var{protocol} to
\class{IOpenProtocol}, and then register with \method{addImplicationListener()}
to receive notice of any protocols that might be implied by \var{protocol}
in future.
\end{methoddesc}

\end{classdesc*}

















\begin{classdesc*}{IImplicationListener}
This interface documents the behavior required of an object supplied to
the \method{IOpenProtocol.addImplicationListener()} method.  Such objects must
be weak-referenceable, usable as a dictionary key, and supply the following
method:

\begin{methoddesc}{newProtocolImplied}{srcProto, destProto, adapter, depth}
Receive notice that an adaptation was declared from \var{srcProto} to
\var{destProto}, using the \class{IAdapterFactory} \var{adapter}, at
implication level \var{depth}.

When used as part of an \class{IOpenProvider} implementation, this method is
typically used to recursively invoke \function{declareAdapterForObject()} with
increased depth and appropriately composed adapters from protocols already
supported by the object.
\end{methoddesc}

\end{classdesc*}























\begin{classdesc*}{IOpenImplementor}
If an class or type supplied to \function{declareAdapterForType} supports
this interface, it will be notified of the declaration and any future
declarations that affect the class, due to current or future protocol
implication relationships.  Supporting this interface is not necessary; it is
provided as a hook for advanced users.  Note that to declare a class or type
as an \class{IOpenImplementor}, you must call \code{adviseObject(theClass,
provides=[IOpenImplementor])} after the class definition or place
\code{advise(classProvides=[IOpenImplementor])} in the body of the class, since
this interface must be provided by the class itself, not by its instances.  (Of
course, if you implement this interface via a metaclass, you can declare that
the metaclass' instances provide the interface.)

Notification to classes supporting \class{IOpenImplementor} occurs via the
following method:

\begin{methoddesc}{declareClassImplements}{protocol, adapter, depth}
Receive notice that instances of the class support \var{protocol} via the
the \class{IAdapterFactory} supplied in \var{adapter}, at implication level
\var{depth}.
\end{methoddesc}

\end{classdesc*}


















\begin{classdesc*}{IAdapterFactory}
An interface documenting the requirements for an object to be used as an
adapter factory: i.e., that it be a callable accepting an object and a
protocol.)  This interface is not used by the \module{protocols} package except
as documentation.
\end{classdesc*}


\begin{classdesc*}{IProtocol}
An interface documenting the basic requirements for an object to be used as
a protocol for \function{adapt()}: i.e., that it be usable as a dictionary
key.  This interface is not used by the \module{protocols} package except as
documentation.
\end{classdesc*}


\begin{classdesc*}{IAdaptingProtocol}
An interface documenting the requirements for a protocol object to be able to
adapt objects when used with \function{adapt()}: i.e., that it have a
\method{__adapt__} method that accepts the object to be adapted and returns
either an object providing the protocol or \constant{None}.  This interface is
not used by the \module{protocols} package except as documentation.  It is a
subclass of \class{IProtocol}, so any implementation of this interface must
support the requirements defined by \class{IProtocol} as well.
\end{classdesc*}
















\subsubsection{\module{protocols.adapters} --- ``Adapter arithmetic'' support \label{protocol-adapters-module}}
\declaremodule{protocols-adapters}{protocols.adapters}

The \module{protocols.adapters} module provides support for doing ``adapter
arithmetic'' such as determining which of two adapter paths is shorter,
composing a new adapter from two existing adapters, and updating an adapter
registry with a new adapter path.  See section \ref{proto-implication} for a
more general discussion of adapter arithmetic.

\begin{funcdesc}{minimumAdapter}{a1, a2 \optional{, d1=0, d2=0}}
Find the ``shortest'' adapter path, \var{a1} at depth \var{d1}, or \var{a2} at
depth \var{d2}.  Assuming \var{a1} and \var{a2} are adapter factories that
accept similar input and return similar output, this function returns the one
which is the ``shortest path'' between its input and its output.  That is, the
one with the smallest implication depth (\var{d1} or \var{d2}), or, if the
depths are equal, then the adapter factory that is composed of the fewest
chained factories (as composed by \function{composeAdapters()}) is returned.
If neither factory is composed of multiple factories, or they are composed of
the same number of intermediate adapter factories, then the following
preference order is used:

\begin{enumerate}
\item If one of the adapters is \function{NO_ADAPTER_NEEDED}, it is returned
\item If one of the adapters is \function{DOES_NOT_SUPPORT}, the \emph{other}
adapter is returned.
\item If both adapters are the exact same object (i.e. \code{a1 is a2}), either
one is returned
\end{enumerate}

If none of the above conditions apply, then the adapter precedence is
considered ambiguous, and a \exception{TypeError} is raised.

This function is used by \function{updateWithSimplestAdapter} to determine
whether a new adapter declaration should result in a registry update.  Note
that the determination of adapter composition length uses the
\member{__adapterCount__} attribute, if present.  (It is assumed to be
\constant{1} if not present.  See \function{composeAdapters()} for more
details.)
\end{funcdesc}


\begin{funcdesc}{composeAdapters}{baseAdapter, baseProtocol, extendingAdapter}
Return a new \class{IAdapterFactory} composed of the input adapter factories
\var{baseAdapter} and \var{extendingAdapter}.  If either input adapter is
\function{DOES_NOT_SUPPORT}, \function{DOES_NOT_SUPPORT} is returned.  If
either input adapter is \function{NO_ADAPTER_NEEDED}, the other input adapter
is returned.  Otherwise, a new adapter factory is created that will return
\code{\var{extendingAdapter}(\var{baseAdapter}(object,\var{baseProtocol}),protocol)}
when called with an \code{object, protocol} pair.  (Note: the actual
implementation verifies that \var{baseAdapter} didn't return \constant{None}
before it calls \var{extendingAdapter}).

If this function creates a new adapter factory, the factory will have an
\member{__adapterCount__} attribute set to the sum of the
\member{__adapterCount__} attributes of the input adapter factories.  If an
input factory does not have an \member{__adapterCount__} attribute, it is
assumed to equal \constant{1}.  This is done so that the
\function{minimumAdapter()} can compare the length of composed adapter chains.
\end{funcdesc}


\begin{funcdesc}{updateWithSimplestAdapter}{mapping, key, adapter, depth}
Treat \var{mapping} as an adapter registry, replacing the entry designated by
\var{key} with an \code{(\var{adapter},\var{depth})} tuple, if and only if
the new entry would be a ``shorter path'' than the existing entry, if any.
(I.e., if \code{minimumAdapter(old, \var{adapter}, oldDepth, \var{depth})}
returns \var{adapter}, and \var{adapter} is not the existing registered
adapter.  The function returns a true value if it updates the contents of
\var{mapping}.

This function is used to manage type-to-protocol, protocol-to-protocol, and
object-to-protocol adapter registries, keyed by type or protocol.  The
\var{mapping} argument must be a mapping providing \method{__setitem__()}
and \method{get()} methods.  Values stored in the mapping will be
\code{(\var{adapter},\var{depth})} tuples.
\end{funcdesc}






\subsubsection{\module{protocols.advice} --- Metaclasses and other ``Magic''}
\declaremodule{protocols-advice}{protocols.advice}
addClassAdvisor

isClassAdvisor

getFrameInfo

getMRO

classicMRO

determineMetaclass

minimalBases


\subsubsection{\module{protocols.classic} --- Adapters to support built-in types
and Zope Interfaces}
\declaremodule{protocols-classic}{protocols.classic}





















\subsection{Customizing and Extending the Declaration API\label{customizing-adaptation}}


\subsubsection{The \class{IOpenProtocol} Interface \label{open-protocols}}

\begin{classdesc*}{IOpenProtocol}
\class{IOpenProtocol} is an interface object representing the behavior a
protocol object must provide, to be usable with the \module{protocols}
declaration API.  (That is, functions whose names begin with \code{declare}
or \code{advise}.)

Although the \module{protocols} package supplies a default implementation of
this interface (\class{protocols.Protocol}), you may use any object that
supports this interface as a protocol for the declaration APIs.

\class{IOpenProtocol} is a subclass of \class{IAdaptingProtocol}.  A complete
description of the methods this interface requires is in section
\ref{protocols-implementing}.
\end{classdesc*}

IOpenProtocol

metaclass issues, fwd ref to metamethod and supermeta


















\subsubsection{Customizing Class/Type Declaration Management}

IOpenImplementor

\subsubsection{Customizing Single-Object Declaration Management}

IOpenProvider

\subsubsection{Using Protocol Objects}
Although most common uses for protocols can be served by using \class{Interface}
objects, there are sometimes situations that call for a dynamically created
protocol, or a new kind of protocol with added or changed behaviors.

For these situations, one may wish to use the \class{Protocol} or
\class{InterfaceClass} classes directly.

Protocol

InterfaceClass

IAdaptingProtocol

IOpenProtocol


















\subsection{Examples}
\subsubsection{Replacing introspection with Adaptation \label{introspect-elim}}







































